\chapter{Systems of ODEs}

\section{First-order systems of ODEs}
Consider the following:
\begin{equation*}
\ddot{x} + x \dot{x} + \sin x = 0
\end{equation*}

This is a 2nd order, non-linear ODE. We can think of this as a system of ODEs by letting $y = \dot{x}$, so $\dot{y} = \ddot{x}$. Then we can rewrite the above as:
\begin{equation*}
\begin{cases}
\dot{x} = y \\
\dot{y} = -x y - \sin x
\end{cases}
\end{equation*}

While this is a first-order system of ODEs, the solution is now two-dimensional, so we can represent it as a vector:
\begin{equation*}
\begin{bmatrix}
\dot{x} \\
\dot{y}
\end{bmatrix} =
\begin{bmatrix}y \\
-x y - \sin x
\end{bmatrix}
\end{equation*}

We call this a \textbf{phase space} representation of the system. The solution to this system will be a curve in the $xy$-plane, which we can visualize as a trajectory of the system.

\dfn{Linear system of ODEs}{
    A system of ODEs is linear if it can be written in the form:
\begin{equation*}
\dot{u} = 
\begin{bmatrix}
\dot{x}_1 \\
\dot{x}_2 \\
\vdots \\
\dot{x}_n
\end{bmatrix} =
\begin{bmatrix}
a_{11} & a_{12} & \cdots & a_{1n} \\
a_{21} & a_{22} & \cdots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
a_{n1} & a_{n2} & \cdots & a_{nn}
\end{bmatrix}
\begin{bmatrix}x_1 \\
x_2 \\
\vdots \\
x_n
\end{bmatrix}
\end{equation*}
}

Letting the matrix be $A$, we can write a linear system of ODEs as:
\[
\boxed{
\ddot{u} = A u
}
\]

\nt{
    If $A$ was a 1x1 matrix, then this would be a scalar and hence a single ode of the form:
\begin{equation*}
\dot{u} = \lambda u
\end{equation*}
which we solve as:
\begin{equation*}
u(t) = u(0) e^{\lambda t}
\end{equation*}
}

Now, let $A \in \bbR^{m \times n}$, and let $u(t) \in \bbR^n$ be a vector of functions. Then we can write the system of ODEs as:
\begin{equation*}
\dot{u}(t) = A u(t)
\end{equation*}

We guess that the solution is of the form:
\begin{equation*}
u(t) = e^{At} u(0)
\end{equation*}

However, we must define what $e^{At}$ means.

Recall the Taylor series expansion of the exponential function:
\begin{equation*}
e^{\lambda t} = \sum_{k=0}^{\infty} \frac{(\lambda t)^k}{k!}
\end{equation*}

Then we can define the matrix exponential as:
\begin{equation*}
e^{At} = \sum_{k=0}^{\infty} \frac{(At)^k}{k!}
\end{equation*}

We take the first derivative of $e^{At}$:
\begin{equation*}
\frac{d}{dt} e^{At} = \sum_{k=0}^{\infty} \frac{A^k t^{k-1}}{(k-1)!} = A \sum_{k=0}^{\infty} \frac{(At)^k}{k!} = A e^{At}
\end{equation*}

Thus, we have:
\begin{equation*}
u(t) = e^{At} u(0)
\end{equation*}
is indeed a solution to the system of ODEs.

In particular, we found that $\dot{u}(t) = A e^{At} u(0) = A u(t)$, which is exactly the system of ODEs we wanted to solve.

\ex{Trivial example}{
    Let $\dot{u} = Au$ where:
\begin{equation*}
A = \begin{bmatrix}
    a & 0 \\
    0 & d
\end{bmatrix}
\end{equation*}

Then we let:
\begin{equation*}
\begin{bmatrix}
\dot{x} \\
\dot{y}
\end{bmatrix} =
\begin{bmatrix}
    a & 0 \\
    0 & d
\end{bmatrix}
\begin{bmatrix}
x \\
y
\end{bmatrix} =
\begin{bmatrix}
    a x \\
    d y
\end{bmatrix}
\end{equation*}

Then solving each equation separately, we get:
\begin{equation*}
x(t) = x(0) e^{a t}, \quad y(t) = y(0) e^{d t}
\end{equation*}

\begin{equation*}
\Rightarrow u(t) = \begin{bmatrix}
x(0) e^{a t} \\
y(0) e^{d t}
\end{bmatrix} = e^{At} u(0)
\end{equation*}

where 
\begin{equation*}
u(0) = \begin{bmatrix}
x(0) \\
y(0)
\end{bmatrix}
\end{equation*}
}

Consider each of the cases:
\begin{itemize}
\item $a = d < 0$: then the solution decays to zero as $t \to \infty$. Stability is achieved in both cases.
\item $a = d > 0$: then the solution grows exponentially as $t \to \infty$. Stability is never achieved.
\item $a < d < 0$: then the solution decays to zero as $t \to \infty$, but the $y$ component decays faster than the $x$ component. Stability is achieved in both cases.
\item $d < a < 0$: then the solution decays to zero as $t \to \infty$, but the $x$ component decays faster than the $y$ component. Stability is achieved in both cases.
\item $a < 0 < d$: then the solution decays to zero as $t \to \infty$, but the $x$ component decays while the $y$ component grows. Stability is never achieved.
\item $d < 0 < a$: then the solution decays to zero as $t \to \infty$, but the $y$ component decays while the $x$ component grows. Stability is never achieved.
\end{itemize}